---
title: "r-py-cheatsheet"
author: "Yann Say"
engine: knitr
execute: 
  freeze: auto
---
# Reading and writing data

## Create a dataframe
::: {.columns}

::: {.column width="45%"}
```{r}
r_df_example <- data.frame(id = letters[1:6],
numeric = 1:6)
r_df_example
```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}

```{python}
import pandas as pd
py_df_example = pd.DataFrame({'id': list('abcdef'),
                 'numeric': list(range(1, 7))
                })

py_df_example
```
:::

:::

## palmers 
::: {.columns}

::: {.column width="45%"}
```{r}
#| warning: false 
#| message: false

library(palmerpenguins)
library(dplyr)

r_df <- penguins

```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
import seaborn as sns
py_df = sns.load_dataset('penguins')

```
:::

:::

## write csv/xlsx
::: {.columns}

::: {.column width="45%"}
```{r}
r_df |> 
  write.csv("penguins.csv")
```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
py_df.to_csv('penguins.csv')
```
:::

:::

## read csv/xlsx
::: {.columns}
::: {.column width="45%"}
```{r}
r_df_csv <- read.csv("penguins.csv")
```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
py_df_csv =	pd.read_csv("penguins.csv")
```
:::
:::

# Data Exploration

## Columns names
::: {.columns}

::: {.column width="45%"}
```{r}
names(r_df)
```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}

```{python}

py_df.columns
```
:::

:::

## dimensions

::: {.columns}

::: {.column width="45%"}
```{r}
dim(r_df)
```
```{r}
nrow(r_df)
```
```{r}
ncol(r_df)
```

:::
::: {.column width="10%"}
:::
::: {.column width="45%"}

```{python}

py_df.shape
```
```{python}

py_df.shape[0]
```
```{python}

py_df.shape[1]
```
:::

:::

## head
::: {.columns}

::: {.column width="45%"}
```{r}
r_df |> head()
```

:::

::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
py_df.head()
```

:::
:::
## table | tally
::: {.columns}

::: {.column width="45%"}
```{r}
r_df$species |>  table()
```
```{r}
r_df |>  group_by(species) |>  tally()
```

:::

::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
py_df['species'].value_counts()

```

:::
:::
## summary
::: {.columns}

::: {.column width="45%"}
```{r}

```

:::

::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}

```

:::
:::

## structure / info
::: {.columns}

::: {.column width="45%"}
```{r}
r_df |> str()
```

:::

::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
py_df.info()
```

:::
:::



# Data manipulation

## copy of dataframe

How to create 2 independent dataframes. In Python, if you `another_df = original_df`, each modification of `another_df` will be reflected in `original_df`. 

::: {.columns}
::: {.column width="45%"}
```{r}
another_r_df <- r_df
```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
another_py_df = py_df.copy()
```
:::
:::

## filter
::: {.columns}

::: {.column width="45%"}
```{r}
penguins |> 
  filter(species == "Adelie")
```
```{r}
r_df[r_df$species == "Adelie",]
```

:::
::: {.column width="10%"}
:::
::: {.column width="45%"}

```{python}
py_df[py_df["species"] == "Adelie"]

```
:::

:::

## select
::: {.columns}

::: {.column width="45%"}
```{r}
r_df$species
```

```{r}
r_df[["species"]]
```

:::

::: {.column width="10%"}
:::
::: {.column width="45%"}

```{python}
py_df.species
```


```{python}
py_df["species"]
```
```{python}
py_df.filter(["species"])
```

:::
:::


## filter & select
::: {.columns}

::: {.column width="45%"}
```{r}
r_df |> 
  filter(species == "Adelie") |> 
  select("bill_length_mm")
```

```{r}
r_df[r_df$species == "Adelie", "bill_length_mm"]
```

:::

::: {.column width="10%"}
:::
::: {.column width="45%"}

```{python}
py_df.loc[py_df["species"] == "Adelie", "bill_length_mm"] 

```

:::
:::

::: {.columns}
::: {.column width="45%"}
```{r}
r_df[1,1]
```
```{r}
r_df[1:6, 1:6]
```

:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
py_df.iloc[0,0]
```
```{python}
py_df.iloc[:6,:6]
```

:::
:::

## mutate
::: {.columns}
::: {.column width="45%"}
```{r}
r_df_example["new_column"] <- "new value"
r_df_example
```
```{r}
r_df_example <- r_df_example |> 
  mutate(new_column2 = paste(id, numeric))
r_df_example
```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
py_df_example["new_column"] = list(range(6))
py_df_example
```

```{python}
py_df_example["new_column2"] = py_df_example.apply(lambda row: row["id"] + str(row["numeric"]), axis = 1)
py_df_example
```


:::
:::
## summarise
::: {.columns}

::: {.column width="45%"}
```{r}
r_df |> 
  group_by(species) |> 
  summarise(bill_length_mm = mean(bill_length_mm, na.rm = T))
```
:::

::: {.column width="10%"}
:::
::: {.column width="45%"}

```{python}
py_df.groupby('species')['bill_length_mm'].mean().reset_index()

# # Mean dep_delay and arr_delay for each season
# dadel_season = flights.groupby('season')['dep_delay', 'arr_delay'].mean().reset_index()
# print(dadel_season)
# 
# # Mean and std delays by origin
# del_ori = flights.groupby('origin')['total_delay', 'dep_delay', 'arr_delay'].agg(['mean', 'std'])
# print(del_ori)

```

:::
:::

## rename

::: {.columns}
::: {.column width="45%"}
```{r}
r_df |>
  rename(Species = species,
         Island = island)
```
```{r}
rename_r_df <- r_df[,c("species", "island")]
names(rename_r_df) <- c("Species", "Island")
names(rename_r_df)
```

:::
::: {.column width="10%"}
:::
::: {.column width="45%"}

```{python}
py_df_rename1 = py_df.copy()
py_df_rename1.rename(columns={'species': 'Species', 'island': 'Island'}, inplace=True)
py_df_rename1
```
```{python}
py_df_rename2 = py_df[["species","island"]].copy()
py_df_rename2.columns = ["Species", "Island"]
py_df_rename2.columns
```
:::
:::

## arrange

::: {.columns}
::: {.column width="45%"}
```{r}
r_df |> 
  arrange(species) |> 
  head()
```
```{r}
r_df |> 
  arrange(desc(species)) |> 
  head()
```

:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
py_df.sort_values(by = "species").head()
```
```{python}
py_df.sort_values(by = "species", ascending = False).head()
```
:::
:::

# control flows and loops

# function

::: {.columns}
::: {.column width="45%"}
```{r}
my_r_function <- function(number) {
  
  my_square = number * number
  return(my_square)
}

my_r_function(5)
```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
def my_py_function(number):
  my_square = number * number
  return my_square

my_py_function(5)
```
:::
:::

## anonymous functions / lambda

# logics

## not

::: {.columns}
::: {.column width="45%"}
```{r}
!TRUE
```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
not(True)
```
:::
:::

## Missing value
::: {.columns}
::: {.column width="45%"}
```{r}
is.na(r_df$bill_length_mm)
```

```{r}
sum(is.na(r_df$bill_length_mm) )
```

```{r}
sum(!is.na(r_df$bill_length_mm) )
```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
py_df["bill_length_mm"].isnull()
```

```{python}
py_df["bill_length_mm"].isnull().sum()

```

```{python}
py_df["bill_length_mm"].notna().sum()

```
:::
:::

## in and not in

```{r}
### %in% 
### str_detect / grepl
###!%in%
```

```{python}
### in 
###not in
```


# Folders and files management

::: {.columns}
::: {.column width="45%"}
```{r}
list.files()	
```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
import os
os.listdir()

```
:::
:::

## types
::: {.columns}
::: {.column width="45%"}
```{r}
typeof(r_df)
```
:::
::: {.column width="10%"}
:::
::: {.column width="45%"}
```{python}
type(py_df)
```
:::
:::